---
title: "gapminder"
author: "Jenny Lee"
date: "2023-02-18"
output: html_document
---
```{r library}
library(utils)
library(tidyverse)
library(ggplot2)
```

# 1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See https://faculty.washington.edu/otoomet/info201-book/
## r-markdown.html#r-markdown-rstudio-creating.

# 2. (2pt) Load data. How many rows/columns do we have?
```{r read}
gap <- read_delim("gapminder 2.csv")
nrow(gap)
ncol(gap)
```
# 3. (2pt) Print a small sample of data. Does it look OK?
```{r sample}
gap %>%
head(3)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# 1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.
```{r countries}
length(unique(gap$iso3))
length(unique(gap$iso2))
length(unique(gap$name))
```

# 2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso3 -codes. What is going on? Can you find it out?

## (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?

```{r each iso-2}
grouped <- gap %>% 
  group_by(iso2) 
count(grouped)
```
## (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries?
## Hint: two of these entitites are CHANISL and NLD CURACAO.
```{r each iso-3}
grouped2 <- gap %>% 
  group_by(iso3) 
count(grouped2)
```
# 3. (2pt) What is the minimum and maximum year in these data?
```{r min and max year}
min_year <- min(gap$time, na.rm = TRUE)
min_year
max_year <- max(gap$time, na.rm = TRUE)
max_year
```
# 3. CO2 emissions (30pt)
##Next, let’s analyze CO2 emissions.
##1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2 and co2_PC. Which years have most missing data?
```{r each missing}
gap %>% 
  select(iso3, name, iso2) %>%
  summarise(n1 = n_distinct(name), n2 = n_distinct(iso2), n3 = n_distinct(iso3))
```
##2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.
```{r plot}
countries <- c("United States of America", "China", "India", "Austrailia", "Belgium")
co2plot_by_country <- gap %>%
  filter(name %in% countries) %>%
  group_by(name, time) %>%
  summarise(co2_a = sum(co2, na.rm = TRUE), .groups = "drop")
  
ggplot(co2plot_by_country, aes(x = time, y= co2_a , color = name) )+
  geom_line()+
  labs(x = "Year", y = "CO2 Emissions over time")+
  ggtitle("CO2 Emissions per capita by country ")
```


##3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the same countries. What does this figure suggest?
```{r co2 emissions per capita}
countries <- c("United States of America", "China", "India", "Austrailia", "Belgium")
co2_emissions_pc <- gap %>%
  filter(name %in% countries) %>%
  group_by(name, time) %>%
  summarize(all_co2_pc = sum(co2_PC, na.rm = TRUE), groups = "drop")

ggplot(co2_emissions_pc, aes(x = time, y = all_co2_pc, color = name)) +
  geom_line() +
  labs(x = "Year", y = "CO2 Emission per Capita") + ggtitle("CO2 EMission per Capita by Country")

cat("THe United States of America has the most CO2 emissions per capita, however, all countries including the United States of America decreases rapidly.")
```
##4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.
#Note: just compute averages over countries and ignore the fact that countries are of #different size. Hint: Americas 2016 should be 4.80.
```{r  across continents}
continent <- gap %>%
  filter(!is.na(co2_PC)) %>%
  group_by(region, time) %>%
  summarize(avg_co2_pc = mean(co2_PC), .groups = "drop") %>%
  filter(time == 2016)
continent
cat("Europe has the highest average co2vper ccapita.")
```
#5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016. Hint: it should look something along these lines:
```{r co2 emissions per capita barplot}
gap %>%
  filter(!is.na(co2_PC), !is.na(region)) %>%
  filter(time == 1960 | time ==2016) %>%
  group_by(region, time) %>%
  mutate(avg_co2_pc = mean(co2_PC)) %>%
  ggplot(aes(x = region, y = avg_co2_pc, fill = factor(time)))+
  geom_bar(stat = "identity", position = "dodge")
```
#6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).
```{r co2 three largest country}
gap %>%
  group_by(region ,name) %>%
  filter(time == 2016) %>%
  filter(!is.na(region) & !is.na(co2_PC) & !is.na(name)) %>%
  summarize(n = co2_PC) %>%
  filter(rank(desc(n)) <= 3 | rank(n) <= 3) %>%
  arrange(desc(n)) %>%
  arrange(region)
```

# 4 GDP per capita (50pt)
-
#Let’s look at GDP per capita (GDP_PC).
#1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
#1960. Make the point size dependent on the country size, and #color those according to the
#continent. Feel free to adjust the plot in other ways to make #it better. Comment what do you see there.
```{r GDP per capita}
gapminder_1960 <- gap %>% filter(time == 1960)
ggplot(gapminder_1960, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region)) +
  geom_point(alpha = 0.7) +
  scale_size_area(max_size = 12) + 
  scale_color_brewer(palette = "Set1") +
  labs(x = "GDP per capita (USD)", Y = "Life expectancy (years)", title = "GDP per capita vs. Life expectancy (1960)") + 
  theme_minimal()
```
#2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r plot for 2019 data}
gapminder_1960 <- gap %>% filter(time == 2019)
ggplot(gapminder_1960, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region)) +
  geom_point(alpha = 0.7) +
  scale_size_area(max_size = 12) + 
  scale_color_brewer(palette = "Set1") +
  labs(x = "GDP per capita (USD)", Y = "Life expectancy (years)", title = "GDP per capita vs. Life expectancy (1960)") + 
  theme_minimal()
```
#3. (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years?
As GDP per capita increase, lifeExpectancy also increased. There is a radical increasing point at about 15000. 

#4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?
#Note: here as average I mean just average over countries, #ignore the fact that countries are of different size.
```{r average life expectancy}
data_1960 <- gap %>% filter (time == 1960)  %>% filter (!is.na(lifeExpectancy))
mean(data_1960$lifeExpectancy)
data_2019 <- gap %>% filter (time == 2019)  %>% filter (!is.na(lifeExpectancy))
mean(data_2019$lifeExpectancy)
```
#5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.
#Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to #group data by continent when using lag(), otherwise your results will be messed up! See #https://faculty.washington.
#edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.

#6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!
```{r histogram of GDP per capita}
gdp_data <- gap %>% 
  filter(!is.na(GDP_PC), time %in% c(1960, 2019)) 

ggplot(gdp_data, aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(position = "dodge") +
  labs(x = "GDP per capita", y = "Count") +
  ggtitle("GDP per capita in 1960 and 2019")
```

#7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
#counting from top.)
#Hint: check out the function rank()!
#Hint2: 17 for 1960.
```{r ranking of US}
gap %>%
    select(name, time, lifeExpectancy) %>% 
    filter(name != "", !is.na(time), !is.na(lifeExpectancy)) %>%
    filter(time == "1960") %>% 
    mutate(rank = rank(desc(lifeExpectancy))) %>% 
    filter(name == "United States of America")
```
#8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year? Hint: 0.0904 for 1960.
```{r 2019}
gp_2019 <- gap %>% 
  filter(name != "", !is.na(time), !is.na(lifeExpectancy)) %>%
  filter(time == "2019") %>%
  mutate(rank = rank(desc(lifeExpectancy)))
us_2019 <- gp_2019 %>%
  filter(name == "United States of America")
rel_rank <- us_2019$rank / n_distinct(gp_2019$name)
rel_rank
```

